Gaussian Process regression for astronomical time-series

### 2022-08-18 src/tex/ms.tex
adding literature plot

Gaussian Processes (GPs) are a powerful class of statistical models, which allow us to define a probability distribution over random functions. Rather than write down an explicit mathematical formula for the function from which some observations are generated, we model the covariance between pairs of samples from the process, using our physical domain knowledge and/or available data to guide our modelling. GP Regression (GPR) has become increasingly popular in the astronomical community over the last decade. Part of the reason for this uptake is the growing availability and importance of time-domain datasets in astronomy. These systematically contain non-trivial random or unknown signals, whether astrophysical or instrumental, that need to be modelled. In many cases, these are nuisance signals, which we need to marginalise over in order to detect or measure other signals robustly. Sometimes, we are interested in the stochastic behaviour itself, and want to infer its characteristics or predict its behaviour. GPR offers a compelling solution: statistically principled, naturally Bayesian, and extremely flexible, yet mathematically simple. To new users, however, the lack of an explicit functional form for the model can make GPR can seem a little arcane. Furthermore, the computational cost of the method, which scales cubically with the dataset size, can be an obstacle. These factors initially impeded its dissemination in the astronomical community, but have been largely overcome in recent years thanks to the availability of user-friendly, computationally optimised software packages.

@@

First introduced in the geophysical literature as a spatial interpolation method by \citet{kriging}, GPR has since been adopted or re-invented in a wide range of other application domains. GPs were used in simulations in a wide range of astronomical sub-fields \citep[see e.g.][]{1980asfr.symp..159B,1988JGR....9311569C,1997ApJ...483L...1P}, but early mentions of GPs for modelling astronomical datasets \citep[see e.g.][]{1976MitAG..38..192D,1978A&A....70..777V,1991MGeo...16..313J} received limited attention.

To the best of our knowledge, the first use GPR in the refereed astronomical literature was published by \citet{prh92a} in the context of quasar variability, and for a long time this remained its main application domain in astronomy. GPR then gradually appeared in other areas, starting with photometric redshift estimation \citep{2006ApJ...647..102W}, then exoplanet transit observations \citep{2009ApJ...704...51C,2012MNRAS.419.2683G} and radial velocity planet searches \citep{2012MNRAS.419.3147A,2014MNRAS.443.2517H}. Nonetheless, GPR remained relatively niche and few astronomers had heard of it until a few years ago. To illustrate this point, we searched on the NASA Astrophysics Data System (ADS) for articles published in refereeed astronomy and astrophysics journals with the words `Gaussian Process' in the title or abstract. Until 2010, the number of such papers in any given year could be counted on the fingers of one hand. From that point onwards it has risen steadily, and in 2021 it was close to 100.

@@

>>>>

Gaussian Processes (GPs) are a powerful class of statistical models, which allow us to define a probability distribution over random functions. Rather than write down an explicit mathematical formula for the function from which some observations are generated, we model the covariance between pairs of samples from the process, using our physical domain knowledge and/or available data to guide our modelling.
While this may seem abstract, GPs have a wide range of applications from modelling of stochastic physical processes, to high dimensional interpolation and smoothing.
In particular, GP Regression (GPR) has become increasingly popular in the astronomical community over the last decade. Part of the reason for this uptake is the growing availability and importance of time-domain datasets in astronomy. These systematically contain non-trivial random or unknown signals, whether astrophysical or instrumental, that need to be modelled. In many cases, these are nuisance signals, which we need to marginalise over in order to detect or measure other signals robustly. Sometimes, we are interested in the stochastic behaviour itself, and want to infer its characteristics or predict its behaviour. GPR offers a compelling solution: statistically principled, naturally Bayesian, and extremely flexible, yet mathematically simple. To new users, however, the lack of an explicit functional form for the model can make GPR can seem a little arcane. Furthermore, the computational cost of the method, which scales cubically with the dataset size, can be an obstacle. These factors initially impeded its dissemination in the astronomical community, but have been largely overcome in recent years thanks to the availability of user-friendly, computationally optimised software packages.

\dan{Add some summary text here. ``In this review...''}

% In this review, we start with a broad and pedagogical introduction to GP methodology in theory and practice.
% We then work through many examples of how GPs have been used in the astronomical literature.
% Discuss

@@

An early use of GPR, was for spatial interpolation in geophysics \citep{kriging}, and GPR has since been adopted or re-invented in a wide range of other application domains. GPs were used in simulations in a wide range of astronomical sub-fields \citep[see e.g.][]{1980asfr.symp..159B,1988JGR....9311569C,1997ApJ...483L...1P}, but early mentions of GPs for modelling astronomical datasets \citep[see e.g.][]{1976MitAG..38..192D,1978A&A....70..777V,1991MGeo...16..313J} received limited attention.

Perhaps the earliest use of GPR in the refereed astronomical literature that will be familiar to a modern reader was published by \citet{prh92a} in the context of quasar variability, and for a long time this remained its main application domain in astronomy. GPR then gradually appeared in other areas, starting with photometric redshift estimation \citep{2006ApJ...647..102W}, then exoplanet transit observations \citep{2009ApJ...704...51C,2012MNRAS.419.2683G} and radial velocity planet searches \citep{2012MNRAS.419.3147A,2014MNRAS.443.2517H}. Nonetheless, GPR remained relatively niche and few astronomers had heard of it until a few years ago. To illustrate this point, we searched on the NASA Astrophysics Data System (ADS) for articles published in refereeed astronomy and astrophysics journals with the words `Gaussian Process' in the title or abstract. Until 2010, the number of such papers in any given year could be counted on the fingers of one hand. From that point onwards it has risen steadily, and in 2021 it was close to 100.

@@

\begin{figure}[ht]
  \centering
  \script{linterature.ipynb}
  \includegraphics[width=\linewidth]{figures/literature.pdf}
  \caption{A search for.}
  \label{fig:literature}
\end{figure}
